Constraint result to the self-interacting dark matter model without dark energy in cosmology

### 2025-04-07 article/article.tex
update data caculation

   The standard Λ-Cold Dark Matter (ΛCDM) model has been remarkably 
   successful in explaining a wide range of cosmological observations, 
   from the Cosmic Microwave Background (CMB) anisotropies to 
   large-scale structure formation. However, there are also many 
   alternative models that have been proposed to explain the
   accelerated expansion of the Universe, such as the dynamical
   dark energy models and the modified gravity models.

   It has been shown that dark matter self-interaction could distribute 
   to the accelerated expansion without dark energy component. According
   to the Boltzmann formalism, the disequilibrium between dark matter particle
   creation and annihilation processes would creat an effective source term 
   making negative pressure just like the dark energy. \cite{refId0}
   investigated the analytical solutions for the simple interacting dark matter (IDM) 
   model, and find that the effective annihilation term is much smaller than the
   results given by the general methods. The gravitational matter creation
   model is mathematically equivalent to one case of the IDM models.

@@

   \cite{refId0} identified two functional forms for which
   the previous Boltzmann equation can be solved analytically.
   Refering to Appendix B in \cite{refId0}, only one of these two is of interest
   because it provides a ``$\propto a^{-3}$'' dependence of the scale factor,
   which is \begin{equation}
      \Psi(a)=aH(a)R(a)=C_1(n+3)a^nH(a)+\kappa C_1^2a^{2m}.
   \end{equation}
   And the total energy density is
   \begin{equation}
      \rho(a)=C_1a^n+\frac{a^{-3}F(a)}{C_2-\int_1^{a}x^{-3}f(x)F(x)\mathrm{d}x},\label{eq:5}
   \end{equation}
   where $f(a)=-\kappa/[aH(a)],$ and the kernal function $F(a)$ has
   the form \begin{equation}
      F(a)=\exp\left[-2\kappa C_1\int_1^{a}\frac{x^{n-1}}{H(x)}\mathrm{d}x\right].
   \end{equation}
   The first term of Eq.(\ref{eq:5}) is the density corresponding to the
   residual matter creation that results from a possible disequilibrium
   between the particle creation and annihilation processes, while the second
   term can be viewed as the energy density of the self-IDM particles that are
   dominated by the annihilation process.

@@

   If $n=0$, the global density evolution can be transformed as
   \begin{equation}
      \rho(a)=C_1+a^{-3}\frac{e^{-2\kappa C_1(t-t_0)}}{C_2-\kappa Z(t)},
   \end{equation}
   where $Z(t)=\int_{t_0}^ta^{-3}e^{-2\kappa C_1(t'-t_0)}\mathrm{d}t'$(\cite{refId0}).
   Using the usual unit-less $\Omega$-like parameterization, we obtain that
   \begin{equation}
      \left(\frac{H}{H_0}\right)^2=\Omega_{1,0}+\frac{\Omega_{1,0}\Omega_{2,0}a^{-3}e^{-2\kappa C_1(t-t_0)}}
      {\Omega_{1,0}+\kappa C_1\Omega_{2,0}Z(t)},\label{eq:8}
   \end{equation}
   where $\Omega_{1,0}=8\pi GC_1/3H_0^2$ and $\Omega_{2,0}=8\pi G/3H_0^2C_2$,
   which related to $\Omega_{\Lambda}$ and $\Omega_m$ in the $\Lambda$CDM model,
   respectively.
   From Eq.({\ref{eq:2}}), we can also give the mass of the DM particle
   related to the range of $\kappa C_1$(in the unit of Gyr${}^{-1}$)
   \begin{equation}
      M_x=\frac{3.325\times10^{-12}}{\kappa C_1}
      \frac{\langle\sigma u\rangle}{10^{-23}}h^2(1-\Omega_{2,0})\,\text{GeV},\label{eq:9}
   \end{equation}
   where $h\equiv H_0/[100\text{km/s/Mpc}]$.

@@

   If $\kappa=0$, the global density evolution can be written as
   \begin{equation}
      \rho(a)=\mathcal{D}a^{-3}+C_1a^{n},
   \end{equation}
   where $\mathcal{D}=C_2-C-1$. The conditions in which the current 
   model acts as a quintessence cosmology are given by $\mathcal{D}>0,
   C_1>0,$ and $w_{\text{IDM}}=-1-n/3$. This solution is mathematically
   equivalent to that of the gravitational matter creation model of().
   The Hubble flow is now given by
   \begin{equation}
      \left(\frac{H}{H_0}\right)^2=\Omega_{2,0}a^{-3}+\Omega_{1,0}a^{n},
   \end{equation}
   where $\Omega_{2,0}=8\pi G\mathcal{D}/3H_0^2$ and 
   $\Omega_{1,0}=8\pi GC_1/3H_0^2$, respectively.(\cite{refId0})

@@

   To constrain the relevant IDM models (\cite{refId0}), we use the newly revised
   observational $H(z)$ data (OHD)(\cite{PhysRevD.71.123001};
   \cite{Daniel.Stern_2010};\cite{M.Moresco_2012};\cite{Zhang_2014};
   \cite{Moresco_2016};\cite{10.1093/mnras/stx301};\cite{10.1093/mnrasl/slv037};
   \cite{Borghi_2022};\cite{Jiao_2023}), the Pantheon+ set of 1701 SNe Ia 
   (\cite{Scolnic_2022}), the QSO data from \cite{Lusso_2020}, 
   the BAO data from SDSS and DESI 2024.

@@

   The $\chi^2$ function for the BAO data is defined as
   \begin{equation}
      \chi_{\text{BAO}}^2=\sum_i\frac{(D_{\text{th}}/r_{\text{d}}-D_{\text{data}}/r_{\text{d}})^2}{\sigma_i^2},
   \end{equation}
   where $D$ refers to $D_{\text{M}}$, $D_{\text{H}}$, or $D_{\text{V}}$, which ara given as
   \begin{eqnarray}
      D_{\text{M}}(z)&=&c\int_0^z\frac{\text{d}z'}{H(z')},\\
      D_{\text{H}}(z)&=&\frac{c}{H(z)},\\
      D_{\text{V}}(z)&=&\left[zD_{\text{M}}^2(z)D_{\text{H}}(z)\right]^{1/3},
   \end{eqnarray}
   and $r_{\text{d}}$ is the sound horizon at the drag epoch, which is given as
   \begin{equation}
      r_{\text{d}}=\int_{z_{\text{drag}}}^{\infty}\frac{c_s(z')\text{d}z'}{H(z')},
   \end{equation}
   where $c_s$ is the sound speed prior to recombination, which is given as
   \begin{equation}
      c_s(z)=\frac{c}{\sqrt{3\left(1+\frac{3\rho_{\mathrm{b}}(z)}{4\rho_{\gamma}(z)}\right)}}\simeq
      \frac{c}{\sqrt{3\left(1+\frac{3\rho_{\mathrm{b}0}}{4\rho_{\gamma0}}\frac{1}{1+z}\right)}}.
   \end{equation}

@@

   We use the 99\% and 1\% quantiles to determine the upper and lower
   limits of the parameters, and the results are shown in Table \ref{tab:5}.

@@

   We would find that the SNe Ia and QSO combined result is most
   similar to SNe Ia result, what may refer that the QSO cannot
   give an effective constraint.

@@

   \begin{table*}[htbp]
      \caption{The constraint results of Model 1}
      \centering
      \begin{tabular}{lcccccc}
         \hline\hline
         Model 1 & OHD & SNe Ia & SNe Ia + QSO & BAO &
          OHD + SNe Ia & OHD + SNe Ia + BAO \\
         \hline
         $\Omega_{2,0}$ & $0.321_{-0.067}^{+0.055}$ & $0.360_{-0.020}^{+0.020}$
          & $0.359_{-0.020}^{+0.020}$ & $0.301_{-0.013}^{+0.013}$
          & $0.330_{-0.016}^{+0.016}$ & $0.313_{-0.010}^{+0.010}$ \\
         $\log_{10}(\kappa C_1/$Gyr${}^{-1})$ & $<-1.81$ & $<-2.30$ & $<-2.35$
          & $<-2.66$ & $<-2.78$ & $<-3.21$ \\
         $\log_{10}(M_x/$GeV) & $>-10.09$ & $>-9.62$ & $>-9.60$
          & $>-9.25$ & $>-9.15$ & $>-8.70$ \\
         \hline
      \end{tabular}
      \label{tab:5}
   \end{table*}

@@

   \begin{equation}
      Const=\mathcal{F}(-\infty)\equiv\lim_{x\to-\infty}\mathcal{F}(x)=
      \frac{\kappa C_1\ln\kappa C_1}{H_0^2\Omega_{1,0}[-4(\kappa C_1)^2+9H_0^2\Omega_{1,0}]}.
   \end{equation}
   The initial time of today $t_0$ can be easily given as
   \begin{equation}
      t_0=2H_0^2\Omega_{1,0}[\mathcal{F}(-\infty)-\mathcal{F}(-H_0)],
   \end{equation}
   and the redshift $z_{\max}$ can be calculated as
   \begin{equation}
      z_{\max}=\exp\left[2H_0^2\Omega_{1,0}\int_{-\infty}^{-H_0}[\mathcal{F}(-\infty)
      -\mathcal{F}(x)]\mathrm{d}x+H_0t_0\right]-1,
   \end{equation}
   and $z_{\max}$ is also a function of $\Omega_{2,0}, \kappa C_1$ and $H_0$.
   \begin{figure}[htbp]
      \centering
      \includegraphics[width=0.7\textwidth]{zmax.pdf}
      \caption{The redshift $z_{\max}$ as $H_0=70$ km/s/Mpc}
   \end{figure}

>>>>

   The standard Λ-Cold Dark Matter (ΛCDM) model has been successful in 
   explaining the accelerated expansion of the Universe (\cite{Riess_1998}; \cite{Perlmutter_1999}), 
   from the cosmic microwave background (CMB) anisotropies (\cite{Bennett_1996}) to 
   the measurement of baryon acoustic oscillations (BAO; \cite{Eisenstein_2005}). However, there are also many 
   alternative models that have been proposed to explain, such as the dynamical
   dark energy models and the modified gravity models.

   It has been shown that dark matter self-interaction could distribute 
   to the accelerated expansion without dark energy component. According
   to the Boltzmann formalism, the disequilibrium between dark matter particle
   creation and annihilation processes would creat an effective source term 
   making negative pressure just like the dark energy. \cite{Basilakos_2009}
   investigated the analytical solutions for the simple interacting dark matter (IDM) 
   model, and find that the effective annihilation term is much smaller than the
   results given by the general methods. The gravitational matter creation
   model is mathematically equivalent to one case of the IDM models.

@@

   \cite{Basilakos_2009} identified two functional forms for which
   the previous Boltzmann equation can be solved analytically.
   Refering to Appendix B in \cite{Basilakos_2009}, only one of these two is of interest
   because it provides a ``$\propto a^{-3}$'' dependence of the scale factor,
   which is \begin{equation}
      \Psi(a)=aH(a)R(a)=C_1(n+3)a^nH(a)+\kappa C_1^2a^{2m}.
   \end{equation}
   And the total energy density is
   \begin{equation}
      \rho(a)=C_1a^n+\frac{a^{-3}F(a)}{C_2-\int_1^{a}x^{-3}f(x)F(x)\mathrm{d}x},\label{eq:5}
   \end{equation}
   where $f(a)=-\kappa/[aH(a)],$ and the kernal function $F(a)$ has
   the form \begin{equation}
      F(a)=\exp\left[-2\kappa C_1\int_1^{a}\frac{x^{n-1}}{H(x)}\mathrm{d}x\right].
   \end{equation}
   The first term of Eq.(\ref{eq:5}) is the density corresponding to the
   residual matter creation that results from a possible disequilibrium
   between the particle creation and annihilation processes, while the second
   term can be viewed as the energy density of the self-IDM particles that are
   dominated by the annihilation process.

@@

   If $n=0$, the global density evolution can be transformed as
   \begin{equation}
      \rho(a)=C_1+a^{-3}\frac{e^{-2\kappa C_1(t-t_0)}}{C_2-\kappa Z(t)},
   \end{equation}
   where $Z(t)=\int_{t_0}^ta^{-3}e^{-2\kappa C_1(t'-t_0)}\mathrm{d}t'$ (\cite{Basilakos_2009}).
   Using the usual unit-less $\Omega$-like parameterization, we obtain that
   \begin{equation}
      \left(\frac{H}{H_0}\right)^2=\Omega_{1,0}+\frac{\Omega_{1,0}\Omega_{2,0}a^{-3}e^{-2\kappa C_1(t-t_0)}}
      {\Omega_{1,0}+\kappa C_1\Omega_{2,0}Z(t)},\label{eq:8}
   \end{equation}
   where $\Omega_{1,0}=8\pi GC_1/3H_0^2$ and $\Omega_{2,0}=8\pi G/3H_0^2C_2$,
   which related to $\Omega_{\Lambda}$ and $\Omega_m$ in the $\Lambda$CDM model,
   respectively.
   From Eq.({\ref{eq:2}}), we can also give the mass of the DM particle
   related to the range of $\kappa C_1$(in the unit of Gyr${}^{-1}$)
   \begin{equation}
      M_x=\frac{3.325\times10^{-12}}{\kappa C_1}
      \frac{\langle\sigma u\rangle}{10^{-23}}h^2(1-\Omega_{2,0})\,\text{GeV},\label{eq:9}
   \end{equation}
   where $h\equiv H_0/[100\text{km/s/Mpc}]$.

@@

   If $\kappa=0$, the global density evolution can be written as
   \begin{equation}
      \rho(a)=\mathcal{D}a^{-3}+C_1a^{n},
   \end{equation}
   where $\mathcal{D}=C_2-C-1$. The conditions in which the current 
   model acts as a quintessence cosmology are given by $\mathcal{D}>0,
   C_1>0,$ and $w_{\text{IDM}}=-1-n/3$. This solution is mathematically
   equivalent to that of the gravitational matter creation model of().
   The Hubble flow is now given by
   \begin{equation}
      \left(\frac{H}{H_0}\right)^2=\Omega_{2,0}a^{-3}+\Omega_{1,0}a^{n},
   \end{equation}
   where $\Omega_{2,0}=8\pi G\mathcal{D}/3H_0^2$ and 
   $\Omega_{1,0}=8\pi GC_1/3H_0^2$, respectively.(\cite{Basilakos_2009})

@@

   To constrain the relevant IDM models (\cite{Basilakos_2009}), we use the newly revised
   observational $H(z)$ data (OHD)(\cite{PhysRevD.71.123001}; 
   \cite{Daniel.Stern_2010}; \cite{M.Moresco_2012}; \cite{Zhang_2014};
   \cite{Moresco_2016}; \cite{10.1093/mnras/stx301}; \cite{10.1093/mnrasl/slv037}; 
   \cite{Borghi_2022}; \cite{Jiao_2023}), the Pantheon+ set of 1701 SNe Ia 
   (\cite{Scolnic_2022}), the quasar data from \cite{Lusso_2020}, 
   the BAO data from SDSS and DESI 2024.

@@

   The $\chi^2$ function for the BAO data is defined as
   \begin{equation}
      \chi_{\text{BAO}}^2=\sum_i\frac{(D_{\text{th}}/r_{\text{d}}-D_{\text{data}}/r_{\text{d}})^2}{\sigma_i^2},
   \end{equation}
   where $D$ refers to $D_{\text{M}}$, $D_{\text{H}}$, or $D_{\text{V}}$, which ara given as
   \begin{eqnarray}
      D_{\text{M}}(z)&=&c\int_0^z\frac{\text{d}z'}{H(z')},\\
      D_{\text{H}}(z)&=&\frac{c}{H(z)},\\
      D_{\text{V}}(z)&=&\left[zD_{\text{M}}^2(z)D_{\text{H}}(z)\right]^{1/3},
   \end{eqnarray}
   and $r_{\text{d}}$ is the sound horizon at the drag epoch, which is given as
   \begin{equation}
      r_{\text{d}}=\int_{z_{\text{drag}}}^{\infty}\frac{c_s(z')\text{d}z'}{H(z')},
   \end{equation}
   where $c_s$ is the sound speed prior to recombination.

@@

   We use the 97.7\% and 2.3\% quantiles (about $2\sigma$) to determine the upper and lower
   limits of the parameters, and the results are shown in Table \ref{tab:5}.

@@

   We would find that the SNe Ia and QSO combined result is most
   similar to SNe Ia result, what may refer that the QSO cannot
   give an effective constraint. It may be due to the data quality
   at the high redshift.

@@

   \begin{table*}[htbp]
      \caption{The constraint results of Model 1}
      \centering
      \begin{tabular}{lcccccc}
         \hline\hline
         Model 1 & OHD & SNe Ia & SNe Ia + QSO & BAO &
          OHD + SNe Ia & OHD + SNe Ia + BAO \\
         \hline
         $\Omega_{2,0}$ & $0.321_{-0.067}^{+0.055}$ & $0.360_{-0.020}^{+0.020}$
          & $0.359_{-0.020}^{+0.020}$ & $0.301_{-0.013}^{+0.013}$
          & $0.330_{-0.016}^{+0.016}$ & $0.313_{-0.010}^{+0.010}$ \\
         $\log_{10}(\kappa C_1/$Gyr${}^{-1})$ & $<-2.17$ & $<-2.51$ & $<-2.57$
          & $<-3.10$ & $<-2.88$ & $<-3.39$ \\
         $\log_{10}(M_x/$GeV) & $>-9.77$ & $>-9.42$ & $>-9.40$
          & $>-8.84$ & $>-9.03$ & $>-8.52$ \\
         \hline
      \end{tabular}
      \label{tab:5}
   \end{table*}

@@

   \begin{equation}
      Const=\mathcal{F}(-\infty)\equiv\lim_{x\to-\infty}\mathcal{F}(x)=
      \frac{\kappa C_1\ln\kappa C_1}{9H_0^4\Omega_{1,0}^2-4H_0^2\Omega_{1,0}(\kappa C_1)^2}.
   \end{equation}
   The initial time of today $t_0$ can be easily given as
   \begin{equation}
      t_0=2H_0^2\Omega_{1,0}[\mathcal{F}(-\infty)-\mathcal{F}(-H_0)],
   \end{equation}
   and the redshift $z_{\max}$ can be calculated as
   \begin{equation}
      z_{\max}=\exp\left[2H_0^2\Omega_{1,0}\int_{-\infty}^{-H_0}[\mathcal{F}(-\infty)
      -\mathcal{F}(x)]\mathrm{d}x+H_0t_0\right]-1,
   \end{equation}
   and $z_{\max}$ is also a function of $\Omega_{2,0}, \kappa C_1$ and $H_0$.
   \begin{figure}[htbp]
      \centering
      \includegraphics[width=0.7\textwidth]{zmax.pdf}
      \caption{The redshift $z_{\max}$ as $H_0=70$ km/s/Mpc}
   \end{figure}
